---
title: "Direct Estimation"
---

Direct estimates are the first estimation step in an area-level SAE workflow. They show what the survey can say before borrowing strength from covariates or spatial structure.

## Inputs

The direct-estimation step uses the standard files prepared earlier:

| File | Required fields |
|---|---|
| Survey indicator file | `unit_id`, `domain_id`, `indicator`, `weight` |
| Domain-size file | `domain_id`, `domain_size` |

The `indicator` column can be binary, such as `1` for adults who walk weekly and `0` otherwise, or numeric for a continuous outcome.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Situation | Estimator | When to use |
|---|---|---|
| Domain size is known and grossing weight is available | Horvitz-Thompson estimator | Anchors estimates to true domain size and gives design-unbiased levels; required for estimating *counts or totals* |
| Domain size is unknown and/or only analysis weights are available | Hájek estimator | Preferred for *means or proportions* when using normalised analysis weights and no reliable domain size is available |

The **Horvitz-Thompson estimator** calculates the direct estimate $\hat{\theta}_d^{DIR}$ in domain ($d$) as:

$$
\hat{\theta}_d^{DIR} =
\frac{\sum_{i \in s_d} w_{di} y_{di}}{N_d}
$$

The **Hájek Estimator** calculates the direct estimate $\hat{\theta}_d$ in domain ($d$) as:

$$
\hat{\theta}_d^{DIR} =
\frac{\sum_{i \in s_d} w_{di} y_{di}}
{\sum_{i \in s_d} w_{di}}.
$$

where $N_d$ is `domain_size`, $w_{di}$ is `weight`, and $y_{di}$ is `indicator`.

Direct estimates are paired with estimated **sampling variances**, telling us how much uncertainty comes from observing only a sample within each domain. In practice, the variance depends on the estimator, weight scale, and survey design information available.

## Evaluation

Evaluate the direct estimates and variances outputs to make a decision how to proceed:

| Scenario | Next step |
|---|---|
| Most domains have usable sample sizes and uncertainty is acceptable | Direct estimates may be enough for reporting. |
| Many domains have no sample, or invalid variances | Revisit the domain choice or redefine estimand. |
| Variances are acceptable but poor precision  | Proceed to [Frequentist Area-Level SAE](freq-area-sae.qmd) to 'tighten' the estimates and their variances |

(*)Here, "usable variance" means a finite, positive sampling variance for the direct estimate in each domain.

In case **Model-based SAE** is necessary, the output needed will be:

- `domain_id`
- `Direct`, the survey-only estimate
- `Variance`, the estimated sampling variance of `Direct`

## Implementation in R
### Functions
Use [`sae::direct()`](https://search.r-project.org/CRAN/refmans/sae/html/direct.html) to calculate direct estimates by domain. It returns core direct-estimation outputs including domain sample size (`SampSize`), direct estimate (`Direct`), estimated standard deviation (`SD`), and coefficient of variation (`CV`). Square `SD` to obtain the `Variance`.

```{r}
library(sae)

direct_fit <- sae::direct(
  y = survey_indicator$indicator,
  dom = survey_indicator$domain_id,
  sweight = survey_indicator$weight,
  domsize = domain_sizes, 
  replace = FALSE, # without replacement (i.e., no duplicate respondents)
)

direct_results <- direct_fit |>
  dplyr::rename(domain_id = Domain) |>
  dplyr::mutate(Variance = SD^2)
```
::::{.callout-note}
In `sae::direct()`, if `domsize` is provided, the Horvitz-Thompson direct estimator is used; if not provided, the Hajek ratio direct estimator is used.
::::

If the survey includes more detailed design information, use the [`survey`](https://cran.r-project.org/package=survey) package for the direct-estimation step. This gives design-aware domain estimates and uncertainty summaries.

```{r}
library(survey)

design <- survey::svydesign(
  ids = ~psu,                 # primary sampling unit
  strata = ~stratum,          # survey stratum ID
  weights = ~weight,          # survey weight
  data = survey_indicator,    # survey indicator file
)

direct_design_fit <- survey::svyby(
  formula = ~indicator,       
  by = ~domain_id,            
  design = design,            
  FUN = svymean,  # other values: svytotal, svyratio, etc.
  vartype = c("se", "ci"),    # output SE, CI, CV
  na.rm = TRUE
)
```

### Worked example

Lorem Ipsum - share real example
National Survey of Wales
