---
title: "[WIP] Direct Estimation"
---

Direct estimates are the first estimation step in an area-level SAE workflow. They show what the survey can say and how precisely, before applying any model-based estimation, i.e., borrowing strength from covariates or spatial structure.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Situation | Estimator | When to use |
|---|---|---|
| Domain size is known and grossing weight is available | Horvitz-Thompson estimator | Anchors estimates to true domain size and gives design-unbiased levels; required for estimating *counts or totals*, but equally valid for *means or proportions* |
| Domain size is unknown and/or only analysis weights are available | Hájek estimator | Can only be used for estimating *means or proportions* when only using analysis weights and no reliable domain size is available |

The **Horvitz-Thompson estimator** calculates the direct estimate $\hat{\theta}_d^{DIR}$ in domain ($d$) as:

$$
\hat{\theta}_d^{DIR} =
\frac{\sum_{i \in s_d} w_{di} y_{di}}{N_d}
$$

The **Hájek Estimator** (or ratio estimator) calculates the direct estimate $\hat{\theta}_d$ in domain ($d$) as:

$$
\hat{\theta}_d^{DIR} =
\frac{\sum_{i \in s_d} w_{di} y_{di}}
{\sum_{i \in s_d} w_{di}}.
$$

where $N_d$ is `domain_size`, $w_{di}$ is `weight`, and $y_{di}$ is `indicator`.

Direct estimates are paired with estimated **sampling variances**, telling us how much uncertainty comes from observing only a sample within each domain. In practice, the variance depends on the estimator, weight scale, and survey design information available.

## Next Step

Use the direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Most domains have usable sample sizes and acceptable uncertainty | Direct estimates may be enough for reporting. |
| Many domains have no sample, or sampling variances are missing, zero, or unreasonable | Revisit the domain choice, estimand, or input data before modelling. |
| Direct estimates are imprecise and a quick model-based approach is enough | Fit a [Frequentist Area-Level SAE](freq-area-sae.qmd) model to borrow strength from covariates. |
| Direct estimates are imprecise, spatial effects matter, or uncertainty needs to be reported as probabilities | Fit a [Bayesian Area-Level SAE with Spatial](bayesian-area-sae.qmd) model to borrow strength from covariates and neighbouring areas. |

## Code Implementation

:::{.callout-tip}
Use [`sae::direct()`](https://search.r-project.org/CRAN/refmans/sae/html/direct.html) to calculate direct estimates by domain. It returns core direct-estimation outputs including domain sample size (`SampSize`), direct estimate (`Direct`), estimated standard deviation (`SD`), and coefficient of variation (`CV`). Square `SD` to obtain the `Variance`.

In `sae::direct()`, if `domsize` is provided, the Horvitz-Thompson direct estimator is used; if not provided, the Hajek ratio direct estimator is used.
:::

It's worth reminding ourselves of the estimands of interest defined at the start first

| Indicator |  Target domain | Domain size | Weight |
|---|---|---|---|
| `commute_distance` |  MSOA |  Worker population by MSOA |`SampleAdultWeight`, analysis weight |
| `poor_bus_satisfaction` |  MSOA | Adult population by MSOA | `WalesAdultWeight` grossing weight |
| `walk_weekly` |  Local Authority |  Total population by Local Authority |`WalesTravelWeight`, grossing weight |

We will perform Direct Estimation for each of them differently according to the estimands and data availability (weight and domain size)

- `commute_distance` -> no domain sizes / only analysis weights -> Hajék ratio estimator
- `poor_bus_satisfaction` -> domain sizes / grossing weights -> Hórvitz-Thompson estimator
- `walk_weekly` -> domain sizes / grossing weights -> Hórvitz-Thompson estimator

```{r}
# Download pre-generated analysis-ready files into a temp folder
library(readr)

zip_tmp <- tempfile(fileext = ".zip")
url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/assets/tutorial-clean-data.zip"
download.file(url, zip_tmp, mode = "wb")
unzip(zip_tmp, exdir = tempdir())
base_path <- file.path(tempdir(), "clean")
```

#### Mean Commute Distance
Load the `commute_distance` survey indicator. It has no matching domain size file.
```{r}
survey_indicator <- read_csv(
  file.path(base_path,"ind_weight_poor_bus_satisfaction.csv")
)

domain_sizes <- read_csv(
  file.path(base_path, "domain_sizes_adult_msoa.csv")
)
```

#### Dissatisfaction with bus service
Load the `poor_bus_satisfaction` survey indicator and its matching adult MSOA domain-size file.


#### Walking
Load the `walk_weekly` survey indicator and its matching total population LA domain-size file.
