---
title: "Direct Estimation"
execute:
  eval: true
---

This workbook focuses on `walk_weekly`. We first produce the direct estimates, then evaluate whether those direct estimates are complete and precise enough to report. The same direct-estimation checks are then repeated for the other indicators, but their code is hidden so the page stays centred on the worked example.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Scenario | Estimator | Formula | When to use |
|---|---|---|---|
| Known Domain Sizes; has grossing weight | Horvitz-Thompson estimator | $\hat{\theta}_d^{DIR}=\frac{\sum_{i\in s_d}w^G_{di}y_{di}}{N_d}$| Anchors estimates to true domain size and gives design-unbiased levels; required for estimating *counts or totals*, but equally valid for *means or proportions* |
| Unknown Domain Sizes; only analysis weights | Hájek estimator |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}w_{di}^Ay_{di}}{\sum_{i \in s_d}w^A_{di}}$| Can only be used for estimating *means or proportions* when only using analysis weights and no reliable domain size is available |
| Unknown Domain Sizes; no survey weights provided | Unweight sample mean |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}y_{di}}{n_d}$| The simplest form of direct estimator, likely biased if survey sampling is not random, and imprecise when $n_d$ is too low  |

Here, the **direct estimate $\hat{\theta}_d^{DIR}$** is for **domain $d$** with **sample size $n_d$** and **domain size $N_d$**, if known.

Direct estimates are paired with estimated **sampling variances**, telling us how much uncertainty comes from observing only a sample within each domain. In practice, the variance depends on the estimator, weight scale, and survey design information available.

## Next Step

Use the direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Most domains have usable sample sizes and acceptable uncertainty | Direct estimates may be enough for reporting. |
| Many domains have no sample, or sampling variances are missing, zero, or unreasonable | Revisit the domain choice, estimand, or input data before modelling. |
| Direct estimates are imprecise, spatial effects matter, or uncertainty needs to be reported as probabilities | Fit a [Bayesian Area-Level SAE with Spatial](bayesian-area-sae.qmd) model to borrow strength from covariates and neighbouring areas. |

## Code Implementation

:::{.callout-tip}
Use [`sae::direct()`](https://search.r-project.org/CRAN/refmans/sae/html/direct.html) to calculate direct estimates by domain. It selects an estimation method automatically based on the provided inputs: 

- Horvitz-Thompson estimator: pass `sweight` and pre-aggregated `domsize`.
- Hajek ratio estimator: pass `sweight`, and `domsize` equals sum of weights as pseudo-domain-size.
- If neither is provided, it defaults to unweighted sample mean.
:::

```{r}
# Download pre-generated analysis-ready files into a temp folder
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sae)
library(sf)
library(ggplot2)
library(patchwork)

# Load analysis-ready data
zip_tmp <- tempfile(fileext = ".zip")
url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-clean-data.zip"
download.file(url, zip_tmp, mode = "wb")
unzip(zip_tmp, exdir = tempdir())
base_path <- file.path(tempdir(), "clean")

# Load shapefiles for plotting
boundary_zip <- tempfile(fileext = ".zip")
boundary_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-boundaries.zip"
download.file(boundary_url, boundary_zip, mode = "wb")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

la_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-ltla.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = LAD21CD) |>
  filter(str_starts(domain_id, "W"))

# Create master lists of MSOAs and LAs
msoa_master_list <- read_csv(
  file.path(base_path, "domain_sizes_adult_msoa.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_name) |>
  distinct()

la_master_list <- read_csv(
  file.path(base_path, "domain_sizes_population_la.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_name) |>
  distinct()
```

### Frequent walking behaviour

`walk_weekly` estimates the proportion of people who walk at least weekly by Local Authority. 

Load the `walk_weekly` survey indicator and its matching total population LA domain-size file.

```{r}
survey_indicator_walk <- read_csv(
  file.path(base_path, "ind_weight_walk_weekly.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id))

domain_sizes_walk <- read_csv(
  file.path(base_path, "domain_sizes_population_la.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_size) |>
  as.data.frame()
```

The indicator has matching domain sizes (i.e., total population by LA), so we use a Horvitz-Thompson direct estimator, simply by specifying `domsize` in the `sae::direct()` function call

```{r}
dir_est_walk <- sae::direct(
  y = survey_indicator_walk$indicator,
  dom = survey_indicator_walk$domain_id,
  sweight = survey_indicator_walk$weight,
  domsize = domain_sizes_walk
)

dir_est_walk$Var <- dir_est_walk$SD^2
dir_est_walk$Lower <- dir_est_walk$Direct - 1.96 * dir_est_walk$SD
dir_est_walk$Upper <- dir_est_walk$Direct + 1.96 * dir_est_walk$SD

head(dir_est_walk)
```


### Mean commute distance

The indicator `commute_distance` has no known worker domain-size file, so the sum of analysis weights in each sampled MSOA is used as a pseudo domain size because `sae::direct()` requires a `domsize`. This is a common workaround when working with `sae` and the Hajek estimator is called for.

```{r}
survey_indicator <- read_csv(
  file.path(base_path, "ind_weight_commute_distance.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id))

domain_sizes_pseudo <- survey_indicator |>
  group_by(domain_id) |>
  summarise(domain_size = sum(weight), .groups = "drop") |>
  as.data.frame()

dir_est_commute <- sae::direct(
  y = survey_indicator$indicator,
  dom = survey_indicator$domain_id,
  sweight = survey_indicator$weight,
  domsize = domain_sizes_pseudo,
  replace = TRUE
)

dir_est_commute$Var <- dir_est_commute$SD^2
dir_est_commute$Lower <- dir_est_commute$Direct - 1.96 * dir_est_commute$SD
dir_est_commute$Upper <- dir_est_commute$Direct + 1.96 * dir_est_commute$SD

head(dir_est_commute)
```


### Dissatisfaction with bus service

The indicator `poor_bus_satisfaction` uses the adult MSOA domain-size file, so it is also estimated with a Horvitz-Thompson direct estimator.

```{r}
survey_indicator_bus <- read_csv(
  file.path(base_path, "ind_weight_poor_bus_satisfaction.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id))

msoa_adult_domain_sizes <- read_csv(
  file.path(base_path, "domain_sizes_adult_msoa.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_size) |>
  as.data.frame()

dir_est_bus <- sae::direct(
  y = survey_indicator_bus$indicator,
  dom = survey_indicator_bus$domain_id,
  sweight = survey_indicator_bus$weight,
  domsize = msoa_adult_domain_sizes
)

dir_est_bus$Var <- dir_est_bus$SD^2
dir_est_bus$Lower <- dir_est_bus$Direct - 1.96 * dir_est_bus$SD
dir_est_bus$Upper <- dir_est_bus$Direct + 1.96 * dir_est_bus$SD

head(dir_est_bus)
```

### Evaluation

Direct estimation itself is prone to be insufficient: some domains may have no sampled observations at all especially when estimating across very small areas, while others may have so few data points that their estimates are not reliable.

To assess the quality of the direct estimates, first check whether any domains lack a direct estimate because they have no samples.

Next, assess whether the available direct estimates are precise enough to report. Two uncertainty metrics are especially important:

- `median_cv`:  the typical coefficient of variation, calculated as the standard error divided by the estimate. Values above 100% indicate that the typical standard error is larger than the estimate itself. 
- `median_ci_width`: the typical width of the 95% confidence interval, expressed in the original units.

We define custom functions to summarise and plot these diagnostics.

```{r}
summarise_direct_estimates <- function(result, master_domains) {
  domains_no_data <- master_domains |>
    anti_join(
      result |> transmute(domain_id = Domain),
      by = "domain_id"
    )
  
  result |>
    summarise(
      target_domain_n = nrow(master_domains),
      domains_with_direct_estimate = n(),
      domains_with_no_data = nrow(domains_no_data),
      median_sample_size = median(SampSize, na.rm = TRUE),
      median_cv = median(CV, na.rm = TRUE),
      median_ci_width = median(Upper - Lower, na.rm = TRUE)
    ) |>
    pivot_longer(
      cols = everything(),
      names_to = "diagnostic",
      values_to = "value"
    )
}

plot_direct_estimates <- function(result, boundaries) {
  map_data <- boundaries |>
    left_join(
      result,
      by = c("domain_id" = "Domain")
    )
  
  estimate_map <- ggplot(map_data) +
    geom_sf(aes(fill = Direct), colour = "white", linewidth = 0.25) +
    scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
    labs(
      fill = "Estimate",
      title = "Direct estimate"
    ) +
    theme_void()
  
  cv_map <- ggplot(map_data) +
    geom_sf(aes(fill = CV), colour = "white", linewidth = 0.25) +
    scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
    labs(
      fill = "CV",
      title = "Coefficient of Variation"
    ) +
    theme_void()
  
  estimate_map + cv_map
}
```

Summarise and plot the diagnostics of the `walk_weekly` direct estimates.
```{r}
summarise_direct_estimates(dir_est_walk, la_master_list)
```
```{r}
plot_direct_estimates(dir_est_walk,la_boundaries)
```
Summarise and plot the diagnostics of the `commute_distance` direct estimates.
```{r}
summarise_direct_estimates(dir_est_commute, msoa_master_list)
```
```{r}
plot_direct_estimates(dir_est_commute, msoa_boundaries)
```
Summarise and plot the diagnostics of the `poor_bus_satisfaction` direct estimates.
```{r}
summarise_direct_estimates(dir_est_bus, msoa_master_list)
```
```{r}
plot_direct_estimates(dir_est_bus,msoa_boundaries)
```

Overall, the `walk_weekly` and `poor_bus_satisfaction` direct estimates are useful as inputs and are relatively stable because the indicator is estimated at local authority level with known population domain sizes. 

By contrast, the estimates for `commute_distance` rely only on weights and have much higher relative uncertainty. A remedy is to redefine the estimand associated with `commute_distance` so that it applies to a population with known domain sizes, such as all adults, rather than a subgroup whose MSOA totals may be unavailable. However, this would essentially change what is being measured.

In general, the higher the sample size per domain, the more stable the estimates. Moreover, when domain sizes are known, precision is also higher. Since model-based estimation methods work on these direct estimates and variances, having a solid base is essential before moving forward.