---
title: "Direct Estimation"
---

Direct estimates are the first estimation step in an area-level SAE workflow. They show what the survey can say before borrowing strength from covariates or spatial structure.

## Inputs

The direct-estimation step uses the standard files prepared earlier:

| File | Required fields |
|---|---|
| Survey indicator file | `unit_id`, `domain_id`, `indicator`, `weight` |
| Domain-size file | `domain_id`, `domain_size` |

The `indicator` column can be binary, such as `1` for adults who walk weekly and `0` otherwise, or numeric for a continuous outcome.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Situation | Estimator | When to use |
|---|---|---|
| Domain size is known and grossing weight is available | Horvitz-Thompson estimator | Anchors estimates to true domain size and gives design-unbiased levels; required for estimating *counts or totals* |
| Domain size is unknown and/or only analysis weights are available | Hájek estimator | Preferred for *means or proportions* when using normalised analysis weights and no reliable domain size is available |

The **Horvitz-Thompson estimator** calculates the direct estimate $\hat{\theta}_d^{DIR}$ in domain ($d$) as:

$$
\hat{\theta}_d^{DIR} =
\frac{\sum_{i \in s_d} w_{di} y_{di}}{N_d}
$$

The **Hájek Estimator** calculates the direct estimate $\hat{\theta}_d$ in domain ($d$) as:

$$
\hat{\theta}_d^{DIR} =
\frac{\sum_{i \in s_d} w_{di} y_{di}}
{\sum_{i \in s_d} w_{di}}.
$$

where $N_d$ is `domain_size`, $w_{di}$ is `weight`, and $y_{di}$ is `indicator`.

Direct estimates are paired with estimated **sampling variances**, telling us how much uncertainty comes from observing only a sample within each domain. In practice, the variance depends on the estimator, weight scale, and survey design information available.

## Next Step

Use the direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Most domains have usable sample sizes and acceptable uncertainty | Direct estimates may be enough for reporting. |
| Many domains have no sample, or sampling variances are missing, zero, or unreasonable | Revisit the domain choice, estimand, or input data before modelling. |
| Direct estimates are imprecise and a quick model-based approach is enough | Fit a [Frequentist Area-Level SAE](freq-area-sae.qmd) model to borrow strength from covariates. |
| Direct estimates are imprecise, spatial effects matter, or uncertainty needs to be reported as probabilities | Fit a [Bayesian Area-Level SAE with Spatial](bayesian-area-sae.qmd) model to borrow strength from covariates and neighbouring areas. |

## Code Implementation

:::{.callout-tip}
Use [`sae::direct()`](https://search.r-project.org/CRAN/refmans/sae/html/direct.html) to calculate direct estimates by domain. It returns core direct-estimation outputs including domain sample size (`SampSize`), direct estimate (`Direct`), estimated standard deviation (`SD`), and coefficient of variation (`CV`). Square `SD` to obtain the `Variance`.

In `sae::direct()`, if `domsize` is provided, the Horvitz-Thompson direct estimator is used; if not provided, the Hajek ratio direct estimator is used.

If the survey includes more detailed design information, use the [`survey`](https://cran.r-project.org/package=survey) package for the direct-estimation step. This gives design-aware domain estimates and uncertainty summaries.
:::

### Worked Example

For the worked example, we use the pre-generated SAE-ready files created from the data-preparation stage. 

```{r}
library(readr)

sae_ready_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/assets/sae-ready-inputs.zip"
sae_ready_dir <- file.path(tempdir(), "sae-ready-inputs")
sae_ready_zip <- file.path(sae_ready_dir, "sae-ready-inputs.zip")

dir.create(sae_ready_dir, showWarnings = FALSE)
download.file(sae_ready_url, sae_ready_zip, mode = "wb")
unzip(sae_ready_zip, exdir = sae_ready_dir)
```

Load the `poor_bus_satisfaction` survey indicator and its matching adult MSOA domain-size file.

```{r}
survey_indicator <- read_csv(
  file.path(sae_ready_dir, "sae-ready-inputs", "ind_weight_poor_bus_satisfaction.csv")
)

domain_sizes <- read_csv(
  file.path(sae_ready_dir, "sae-ready-inputs", "domain_sizes_adult_msoa.csv")
)
```

Lorem Ipsum - share real example National Survey of Wales
