---
title: "Direct Estimation"
execute:
  eval: true
---

Direct estimation is the process of deriving a preliminary estimate at the desired estimation geography, based solely on the survey data. The direct estimates and their variances are also components in model-based area-level estimation in the next workbook.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Scenario | Estimator | Formula | When to use |
|---|---|---|---|
| Known domain sizes; has grossing weights | Horvitz-Thompson estimator | $\hat{\theta}_d^{DIR}=\frac{\sum_{i\in s_d}w^G_{di}y_{di}}{N_d}$| Anchors estimates to the true population size of each estimation area; required for *counts/totals* and useful for population rates, but equally valid for *means or proportions* |
| Unknown domain sizes; has survey weights | Hájek estimator |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}w_{di}^Ay_{di}}{\sum_{i \in s_d}w^A_{di}}$| Can only be used for estimating *means or proportions* when only using analysis weights and no reliable area population size is available |
| Unknown domain sizes; no survey weights | Unweight sample mean |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}y_{di}}{n_d}$| The simplest form of direct estimator, likely biased if survey sampling is not random, and imprecise when $n_d$ is too low  |

Here, the **direct estimate $\hat{\theta}_d^{DIR}$** is for **area $d$** with **sample size $n_d$** and **domain size $N_d$** if known. 

Direct estimates are paired with **sampling variances**, telling us how much uncertainty comes from observing only a sample within each estimation area. The variance depends not only on the estimator, but also the weight scale, and survey design information available.

## Evaluate direct estimates

Use the direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Too many areas have no/few samples, variances are missing or unreasonably large | Revisit the estimand definition (target population, estimation geography) |
| All areas have direct estimates with acceptable sample sizes, and uncertainty | Direct estimates may be enough for reporting. |
| Most areas have direct estimates, albeit imprecise. Spatial effects matter. | Fit a [Area-Level Estimation Model](bayesian-area-sae.qmd) model to borrow strength from covariates and neighbouring areas. |

## Pratical: Code Implementation

:::{.callout-tip}
Use `survey::svyby()` for flexible calculation of direct estimates using any estimator and input survey design elements such as strata.
:::

### Setup
```{r}
# Load libraries
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)

# Load analysis-ready data into a temp folder
zip_tmp <- tempfile(fileext = ".zip")
url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-clean-data.zip"
download.file(url, zip_tmp, mode = "wb")
unzip(zip_tmp, exdir = tempdir())
base_path <- file.path(tempdir(), "clean")

# Load shapefiles for plotting
boundary_zip <- tempfile(fileext = ".zip")
boundary_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-boundaries.zip"
download.file(boundary_url, boundary_zip, mode = "wb")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

# List of all MSOAs for summary
msoa_domains <- msoa_boundaries |>
  st_drop_geometry() |>
  select(domain_id)
```

### Frequent walkers per 1,000 residents by MSOA
#### Direct Estimation

Keep in mind that with `freq_walker`, we are estimating:

> **Rate per 1,000 residents** walking as main transport mode daily across **MSOAs**

`freq_walker` is a binary survey indicator for whether a respondent walks for transport daily. Our estimand is a rate estimand (per 1k), so we need the HT direct estimator approach. More specifically, the estimate is equal to a population-weighted total of frequent walkers per MSOA, divided by total population in each MSOA (HT estimator). The ratio is then multiplied by 1000 to represent rate per 1k.

- For the weighted total in the numerator, we multiple the binary indicator with its companion grossing weight. 
- For domain sizes in the denominator, i.e., total population per MSOA, we can simply sum the grossing `weight` within each MSOA. This is appropriate because the grossing weight we have was explicitly scaled from the analysis weight using the MSOA domain sizes in the previous workbook. In practice, some surveys provide grossing weights directly, and these may not sum to the target denominator needed. In that case, use the actual domain sizes as the denominator instead.

```{r}
# Load the active-travel indicator with its grossing weight
survey_indicator_walk <- read_csv(
  file.path(base_path, "ind_weight_freq_walker.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator),
    !is.na(weight),
    !is.na(domain_id),
    !is.na(strata)
  )

# Calculate sum of grossing weight to be used as domain sizes (denominator)
survey_indicator_walk <- survey_indicator_walk |>
  group_by(domain_id) |>
  mutate(domain_size = sum(weight, na.rm = TRUE)) |>
  ungroup()

# Set up the survey design
design_walk <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_walk
)

# Estimate the weighted total in each MSOA (numerator)
walk_total <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id + domain_size,
  design = design_walk,
  FUN = survey::svytotal,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Divide by the denominator and keep the uncertainty on the same scale
dir_est_walk <- walk_total |>
  transmute(
    domain_id,
    direct = (indicator / domain_size) * 1000,
    variance = ((se / domain_size) * 1000)^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_walk)
```
Alongside the direct estimate (`direct`) and its sampling variance (`variance`), the output above also produces 

- `cv` for the coefficient of variation calculated as the standard error divided by the estimate. Values >1 indicate that the typical SE is larger than the estimate.
- `ci_width` for the approximate width of the 95% confidence interval (expressed in the original units).

#### Evaluation
Direct estimation itself is prone to be insufficient. To assess the quality of the direct estimates, 

- First check whether any estimation areas lack a direct estimate. In some cases, they might have a direct estimate but no useful variance, especially when `n = 1`. 
- Next, assess whether the available direct estimates are precise enough to report by looking at metrics such as `median_cv` and `median_ci_width`.

```{r}
# Plot estimate and CV
walk_map_data <- msoa_boundaries |>
  left_join(dir_est_walk, by = "domain_id")

walk_estimate_map <- ggplot(walk_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
  labs(
    fill = "Estimate",
    title = "Direct estimate"
  ) +
  theme_void()

walk_cv_map <- ggplot(walk_map_data) +
  geom_sf(aes(fill = cv), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
  labs(
    fill = "CV",
    title = "Coefficient of Variation"
  ) +
  theme_void()

walk_estimate_map + walk_cv_map +
  plot_annotation(
    title = "Residents walking daily as main transport mode per 1,000 residents by MSOA"
  )
```

```{r}
# Uncertainty diagnostics summary
walk_sample_n <- survey_indicator_walk |>
  count(domain_id, name = "n")

msoa_domains |>
  left_join(walk_sample_n, by = "domain_id") |>
  left_join(dir_est_walk, by = "domain_id") |>
  mutate(n = replace_na(n, 0)) |>
  summarise(
    areas_total = n(),
    areas_n_0 = sum(n == 0),
    areas_n_lt2 = sum(n < 2),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

The direct direct estimates appear weak with the maps show many areas with high uncertainty, The diagnostics confirm that this is driven by sparse subsampled active-travel module data: several MSOAs have no usable observations, more have fewer than two observations, and the median CV is high. This makes it a strong candidate for model-based estimation, where covariates and spatial structure can help stabilise estimates for areas with little direct survey evidence.

### Mean commute distance among commuters by MSOA
#### Direct Estimation
The indicator `commute_distance` is continuous, so the direct estimate is simply the weighted mean commute distance among available commuter respondents (Hajek estimator).

```{r}
# Load the indicator and weight
survey_indicator_commute <- read_csv(
  file.path(base_path, "ind_weight_commute_distance.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator), 
    !is.na(weight), 
    !is.na(domain_id), 
    !is.na(strata)
    )

# Set up the survey design
design_commute <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_commute
)

# Estimate the weighted mean in each MSOA
commute_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_commute,
  FUN = survey::svymean,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Output with uncertainty measures
dir_est_commute <- commute_mean |>
  transmute(
    domain_id,
    direct = indicator,
    variance = se^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_commute)
```
#### Evaluation
The codes to summarise and plot direct estimates and uncertainity measures mirror the ones above and have been hidden.
```{r}
#| echo: false
commute_map_data <- msoa_boundaries |>
  left_join(
    dir_est_commute,
    by = "domain_id"
  )

commute_estimate_map <- ggplot(commute_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
  labs(
    fill = "Estimate",
    title = "Direct estimate"
  ) +
  theme_void()

commute_cv_map <- ggplot(commute_map_data) +
  geom_sf(aes(fill = cv), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
  labs(
    fill = "CV",
    title = "Coefficient of Variation"
  ) +
  theme_void()

commute_estimate_map + commute_cv_map +
  plot_annotation(
    title = "Mean Commute Distance among commuters by MSOA"
  )
```

```{r}
#| echo: false
commute_sample_n <- survey_indicator_commute |>
  count(domain_id, name = "n")

msoa_domains |>
  left_join(commute_sample_n, by = "domain_id") |>
  left_join(dir_est_commute,by = "domain_id") |>
  mutate(n = replace_na(n, 0)) |>
  summarise(
    areas_total = n(),
    areas_n_0 = sum(n == 0),
    areas_n_lt2 = sum(n < 2),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

The direct estimates are more stable, where most MSOAs have at least some usable commuter observations, and the median sample size is higher. Even so, the confidence intervals remain fairly wide in practical terms, so the direct estimates should still be treated as noisy area-level summaries rather than precise local values.

### Dissatisfied bus users among recent adult bus riders by MSOA

#### Direct Estimation

The indicator `bus_nosat` is binary. It is defined only among adults who rode a bus in the last 12 months: `1` means dissatisfied, `0` means not dissatisfied, and non-riders are left as nulls. The direct estimate is the weighted proportion among available recent bus riders (Hajek estimator).

```{r}
# Load the indicator and weight
survey_indicator_bus <- read_csv(
  file.path(base_path, "ind_weight_bus_nosat.csv"),
  show_col_types = FALSE
) |>
  filter(
    !is.na(indicator), 
    !is.na(weight), 
    !is.na(domain_id), 
    !is.na(strata)
    )

# Set up the survey design 
design_bus <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_bus
)

# Estimate the weighted share in each MSOA
bus_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_bus,
  FUN = survey::svymean, # when paired with binary indicator, it represents share
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Output with uncertainty measures
dir_est_bus <- bus_mean |>
  transmute(
    domain_id,
    direct = indicator,
    variance = se^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_bus)
```
#### Evaluation
The codes to summarise and plot direct estimates and uncertainity measures mirror the ones above and have been hidden.
```{r}
#| echo: false
bus_map_data <- msoa_boundaries |>
  left_join(
    dir_est_bus,
    by = "domain_id"
  )

bus_estimate_map <- ggplot(bus_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
  labs(
    fill = "Estimate",
    title = "Direct estimate"
  ) +
  theme_void()

bus_cv_map <- ggplot(bus_map_data) +
  geom_sf(aes(fill = cv), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
  labs(
    fill = "CV",
    title = "Coefficient of Variation"
  ) +
  theme_void()

bus_estimate_map + bus_cv_map +
  plot_annotation(
    title = "Prevalence of dissatisfaction with bus services among recent adult bus riders by MSOA"
  )
```

```{r}
#| echo: false
bus_sample_n <- survey_indicator_bus |>
  count(domain_id, name = "n")

msoa_domains |>
  left_join(bus_sample_n, by = "domain_id") |>
  left_join(dir_est_bus,by = "domain_id") |>
  mutate(n = replace_na(n, 0)) |>
  summarise(
    areas_total = n(),
    areas_n_0 = sum(n == 0),
    areas_n_lt2 = sum(n < 2),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

The direct estimates have reasonable geographic coverage, and slightly larger median sample size per MSOA. but they remain imprecise because the estimand is restricted to adults who rode a bus in the last 12 months. This smaller eligible group reduces the effective sample size within many MSOAs, which is reflected in the CV map and summary diagnostics. The direct estimates are useful for diagnosing the pattern of evidence, but model-based estimation is likely to produce more stable local estimates.

### Final thoughts
Even though direct estimation does not involve modelling, it is where we stress-test the raw data. It shows how much survey evidence we actually have in each area, how uncertain the estimates are, and whether the direct estimates and sampling variances are robust enough to feed into the modelled-estimation step in the next workbook.

The **direct estimate calculation** should follow the measure and the estimand:

- For the estimand based on `freq_walker`, the estimate is a **population rate**: the number of frequent walkers divided by the MSOA population, multiplied by 1,000. This requires grossing weights so the survey responses can be scaled to population totals.
- For the estimand based on `commute_distance`, the estimate is the **mean** commute distance among commuter respondents.
- For the estimand based on `bus_nosat`, the estimate is the **share** of recent adult bus riders who are dissatisfied with bus services.

The **uncertainty diagnostics** above show different levels of uncertainty across the three examples. Model-based estimation in the next steps can improve the estimates by reducing uncertainty and producing statistically informed estimates for areas with little or no survey data by borrowing strength from covariates and spatial structure.
