---
title: "Direct Estimation"
execute:
  eval: true
---

Direct estimation is the process of deriving a preliminary estimate at the desired estimation geography, based solely on the survey data. The direct estimates and their variances are also components in model-based area-level estimation in the next workbook.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Scenario | Estimator | Formula | When to use |
|---|---|---|---|
| Known domain sizes; has grossing weights | Horvitz-Thompson estimator | $\hat{\theta}_d^{DIR}=\frac{\sum_{i\in s_d}w^G_{di}y_{di}}{N_d}$| Anchors estimates to the true population size of each estimation area; required for *counts/totals* and useful for population rates, but equally valid for *means or proportions* |
| Unknown domain sizes; has survey weights | Hájek estimator |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}w_{di}^Ay_{di}}{\sum_{i \in s_d}w^A_{di}}$| Can only be used for estimating *means or proportions* when only using analysis weights and no reliable area population size is available |
| Unknown domain sizes; no survey weights | Unweight sample mean |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}y_{di}}{n_d}$| The simplest form of direct estimator, likely biased if survey sampling is not random, and imprecise when $n_d$ is too low  |

Here, the **direct estimate $\hat{\theta}_d^{DIR}$** is for **area $d$** with **sample size $n_d$** and **domain size $N_d$** if known. Direct estimates are paired with estimated **sampling variances**, telling us how much uncertainty comes from observing only a sample within each estimation area. The variance depends not only on the estimator, but also the weight scale, and survey design information available.

## Evaluate direct estimates

Use the direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Too many areas have no/few samples, variances are missing or unreasonably large | Revisit the estimand definition (target population, estimation geography) |
| All areas have direct estimates with acceptable sample sizes, and uncertainty | Direct estimates may be enough for reporting. |
| Most areas have direct estimates, albeit imprecise. Spatial effects matter. | Fit a [Area-Level Estimation Model](bayesian-area-sae.qmd) model to borrow strength from covariates and neighbouring areas. |

## Code Implementation

:::{.callout-tip}
Use `survey::svyby()` for flexible calculation of direct estimates using any estimator and input survey design elements such as strata.
:::

### Setup
```{r}
# Load libraries
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sf)
library(ggplot2)
library(patchwork)

# Load analysis-ready data into a temp folder
zip_tmp <- tempfile(fileext = ".zip")
url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-clean-data.zip"
download.file(url, zip_tmp, mode = "wb")
unzip(zip_tmp, exdir = tempdir())
base_path <- file.path(tempdir(), "clean")

# Load shapefiles for plotting
boundary_zip <- tempfile(fileext = ".zip")
boundary_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-boundaries.zip"
download.file(boundary_url, boundary_zip, mode = "wb")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

msoa_totalpop_domsizes <- read_csv(
  file.path(base_path, "domain_sizes_population_msoa.csv"),
  show_col_types = FALSE
) |>
  mutate(domain_size = as.numeric(domain_size)) |>
  as.data.frame()

```

We will first focus on producing and evaluating the estimates using the indicator `freq_walker` before replicating the process for `bus_nosat` and `commute_distance`.

### Estimand: Frequent walkers per 1,000 residents by MSOA
#### Direct Estimation

Keep in mind that with `walk-weekly`, we are estimating:

> **Rate per 1,000 residents** walking as main transport mode at least weekly across **MSOAs**

`freq_walker` is a binary survey indicator for whether a respondent walks for transport at least weekly. We first scale `SampleTravelWeight` to a Wales-wide grossing weight using the total population across Welsh MSOAs. Then we estimate the weighted total number of weekly walkers in each MSOA, divide by the known MSOA population, and multiply by 1,000.

```{r}
# Load the active-travel indicator with its module analysis weight
survey_indicator_walk <- read_csv(
  file.path(base_path, "ind_weight_freq_walker.csv"),
  show_col_types = FALSE
)

# Scale SampleTravelWeight once to the Wales population total
wales_population_total <- sum(msoa_totalpop_domsizes$domain_size, na.rm = TRUE)
travel_weight_total <- sum(survey_indicator_walk$weight, na.rm = TRUE)
wales_travel_scale <- wales_population_total / travel_weight_total

survey_indicator_walk <- survey_indicator_walk |>
  mutate(WalesTravelWeight = weight * wales_travel_scale)

# Check that the scaled weight sums to the target population
survey_indicator_walk |>
  summarise(
    WalesTravelWeight_sum = sum(WalesTravelWeight, na.rm = TRUE),
    Wales_population_total = wales_population_total
  )
```
```{r}
# Attach the domain size (total pop per MSOA)
survey_indicator_walk <- survey_indicator_walk |>
  left_join(
    msoa_totalpop_domsizes |> select(domain_id, domain_size),
    by = "domain_id"
  )

survey_indicator_walk <- survey_indicator_walk |>
  filter(
    !is.na(indicator),
    !is.na(WalesTravelWeight),
    !is.na(domain_id),
    !is.na(strata),
    !is.na(domain_size)
  )

design_walk <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~WalesTravelWeight,
  data = survey_indicator_walk
)

# Estimate the total number of freq walkers in each MSOA
walk_total <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id + domain_size,
  design = design_walk,
  FUN = survey::svytotal,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

# Convert the total into a rate per 1,000 residents
dir_est_walk <- walk_total |>
  transmute(
    domain = domain_id,
    direct = (indicator / domain_size) * 1000,
    variance = ((se / domain_size) * 1000)^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_walk)
```
Alongside the direct estimate (`direct`) and its sampling variance (`variance`), the output above also produces 

- `cv` for the coefficient of variation calculated as the standard error divided by the estimate. Values >1 indicate that the typical SE is larger than the estimate.
- `ci_width` for the approximate width of the 95% confidence interval (expressed in the original units).

#### Evaluation

Direct estimation itself is prone to be insufficient. For `freq_walker`, the direct estimate depends on only the respondents observed in each MSOA. Small samples can therefore create unstable rates and wide confidence intervals, even though the calculation uses the known MSOA population as the denominator. Since the weight is scaled once for Wales rather than calibrated within each MSOA, some small-sample area estimates may even exceed 1,000 and should be treated as unstable direct estimates.

To assess the quality of the direct estimates, first check whether any estimation areas lack a direct estimate. In some cases, they might have a direct estimate but no useful variance, especially when `n = 1`. Next, assess whether the available direct estimates are precise enough to report by looking at `median_cv` and `median_ci_width`.

```{r}
walk_map_data <- msoa_boundaries |>
  left_join(
    dir_est_walk,
    by = c("domain_id" = "domain")
  )

walk_estimate_map <- ggplot(walk_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
  labs(
    fill = "Estimate",
    title = "Direct estimate"
  ) +
  theme_void()

walk_cv_map <- ggplot(walk_map_data) +
  geom_sf(aes(fill = cv), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
  labs(
    fill = "CV",
    title = "Coefficient of Variation"
  ) +
  theme_void()

walk_estimate_map + walk_cv_map +
  plot_annotation(
    title = "Residents walking weekly as main transport mode per 1,000 residents by MSOA"
  )
```

```{r}
walk_diagnostics <- msoa_totalpop_domsizes |>
  select(domain_id) |>
  left_join(
    survey_indicator_walk |>
      count(domain_id, name = "n"),
    by = "domain_id"
  ) |>
  left_join(
    dir_est_walk,
    by = c("domain_id" = "domain")
  ) |>
  mutate(n = replace_na(n, 0L))

walk_diagnostics |>
  summarise(
    areas_total = nrow(msoa_totalpop_domsizes),
    areas_no_data = sum(n == 0, na.rm = TRUE),
    areas_n_1 = sum(n == 1, na.rm = TRUE),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

### Estimand: Mean commute distance among commuters by MSOA
#### Direct Estimation
The indicator `commute_distance` is continuous, so the direct estimate is the weighted mean commute distance among available commuter respondents.

```{r}
survey_indicator <- read_csv(
  file.path(base_path, "ind_weight_commute_distance.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id), !is.na(strata))

design_commute <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator
)

commute_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_commute,
  FUN = survey::svymean,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

dir_est_commute <- commute_mean |>
  transmute(
    domain = domain_id,
    direct = indicator,
    variance = se^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_commute)
```
#### Evaluation

```{r}
#| echo: false
commute_map_data <- msoa_boundaries |>
  left_join(
    dir_est_commute,
    by = c("domain_id" = "domain")
  )

commute_estimate_map <- ggplot(commute_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
  labs(
    fill = "Estimate",
    title = "Direct estimate"
  ) +
  theme_void()

commute_cv_map <- ggplot(commute_map_data) +
  geom_sf(aes(fill = cv), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
  labs(
    fill = "CV",
    title = "Coefficient of Variation"
  ) +
  theme_void()

commute_estimate_map + commute_cv_map +
  plot_annotation(
    title = "Mean Commute Distance among commuters by MSOA"
  )
```

```{r}
#| echo: false
commute_diagnostics <- msoa_totalpop_domsizes |>
  select(domain_id) |>
  left_join(
    survey_indicator |>
      count(domain_id, name = "n"),
    by = "domain_id"
  ) |>
  left_join(
    dir_est_commute,
    by = c("domain_id" = "domain")
  ) |>
  mutate(n = replace_na(n, 0L))

commute_diagnostics |>
  summarise(
    areas_total = nrow(msoa_totalpop_domsizes),
    areas_no_data = sum(n == 0, na.rm = TRUE),
    areas_n_1 = sum(n == 1, na.rm = TRUE),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

### Estimand: Dissatisfaction with bus service among recent adult bus riders by MSOA

The indicator `bus_nosat` is binary. It is defined only among adults who rode a bus in the last 12 months: `1` means dissatisfied, `0` means not dissatisfied, and non-riders are left as missing. The direct estimate is the weighted proportion among available recent bus riders.

#### Direct Estimation

```{r}
survey_indicator_bus <- read_csv(
  file.path(base_path, "ind_weight_bus_nosat.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id), !is.na(strata))

design_bus <- survey::svydesign(
  ids = ~1,
  strata = ~strata,
  weights = ~weight,
  data = survey_indicator_bus
)

bus_mean <- survey::svyby(
  formula = ~indicator,
  by = ~domain_id,
  design = design_bus,
  FUN = survey::svymean,
  vartype = "se",
  na.rm = TRUE
) |>
  as.data.frame() |>
  tibble::as_tibble()

dir_est_bus <- bus_mean |>
  transmute(
    domain = domain_id,
    direct = indicator,
    variance = se^2,
    cv = if_else(direct == 0, NA_real_, abs(sqrt(variance) / direct)),
    ci_width = 2 * 1.96 * sqrt(variance)
  )

head(dir_est_bus)
```
#### Evaluation

```{r}
#| echo: false
bus_map_data <- msoa_boundaries |>
  left_join(
    dir_est_bus,
    by = c("domain_id" = "domain")
  )

bus_estimate_map <- ggplot(bus_map_data) +
  geom_sf(aes(fill = direct), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
  labs(
    fill = "Estimate",
    title = "Direct estimate"
  ) +
  theme_void()

bus_cv_map <- ggplot(bus_map_data) +
  geom_sf(aes(fill = cv), colour = "white", linewidth = 0.25) +
  scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
  labs(
    fill = "CV",
    title = "Coefficient of Variation"
  ) +
  theme_void()

bus_estimate_map + bus_cv_map +
  plot_annotation(
    title = "Prevalence of dissatisfaction with bus services among recent adult bus riders by MSOA"
  )
```

```{r}
#| echo: false
bus_diagnostics <- msoa_totalpop_domsizes |>
  select(domain_id) |>
  left_join(
    survey_indicator_bus |>
      count(domain_id, name = "n"),
    by = "domain_id"
  ) |>
  left_join(
    dir_est_bus,
    by = c("domain_id" = "domain")
  ) |>
  mutate(n = replace_na(n, 0L))

bus_diagnostics |>
  summarise(
    areas_total = nrow(msoa_totalpop_domsizes),
    areas_no_data = sum(n == 0, na.rm = TRUE),
    areas_n_1 = sum(n == 1, na.rm = TRUE),
    median_sample_size = median(n, na.rm = TRUE),
    median_cv = median(cv, na.rm = TRUE),
    median_ci_width = median(ci_width, na.rm = TRUE)
  ) |>
  pivot_longer(
    cols = everything(),
    names_to = "diagnostic",
    values_to = "value"
  )
```

### Final thoughts
In this exercise, all three direct estimates use the same survey-design pattern: create a stratified survey design, estimate by MSOA, and convert the result to the estimand scale where needed. For `freq_walker`, `SampleTravelWeight` is first scaled to a Wales-wide grossing weight, then the weighted total is divided by the known MSOA population and multiplied by 1,000.

This keeps the code consistent across a mean, a share, and a rate-style binary indicator, and it lets the variance estimates account for the survey strata. The population totals are used as known denominators for `freq_walker`, but the grossing weight is scaled once for Wales overall rather than separately within each MSOA. That is why the diagnostic checks matter: small-sample area estimates can still be noisy even when the denominator is known.

Precision still needs to be judged from the diagnostics rather than from the estimator alone. In general, the larger the sample sizes, the lower the uncertainty. If many areas have no direct estimate, high CVs, or wide confidence intervals, direct estimation is probably not enough. These direct estimates and sampling variances are then better treated as inputs to model-based SAE.
