---
title: "Direct Estimation"
execute:
  eval: true
---

Direct estimates are the first estimation step in an area-level SAE workflow. They show what the survey can say and how precisely, before applying any model-based estimation, i.e., borrowing strength from covariates or spatial structure.

## Direct Estimator and Sampling Variances

The estimator should match the information available in the survey file and the estimand definition.

| Scenario | Estimator | Formula | When to use |
|---|---|---|---|
| Known Domain Sizes; has grossing weight | Horvitz-Thompson estimator | $\hat{\theta}_d^{DIR}=\frac{\sum_{i\in s_d}w^G_{di}y_{di}}{N_d}$| Anchors estimates to true domain size and gives design-unbiased levels; required for estimating *counts or totals*, but equally valid for *means or proportions* |
| Unknown Domain Sizes; only analysis weights | Hájek estimator |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}w_{di}^Ay_{di}}{\sum_{i \in s_d}w^A_{di}}$| Can only be used for estimating *means or proportions* when only using analysis weights and no reliable domain size is available |
| Unknown Domain Sizes; no survey weights provided | Unweight sample mean |$\hat{\theta}_d^{DIR}=\frac{\sum_{i \in s_d}y_{di}}{n_d}$| The simplest form of direct estimator, likely biased if survey sampling is not random and imprecise when $n_d$ is too low  |

Here, the **direct estimate $\hat{\theta}_d^{DIR}$** is for **domain $d$** with **sample size $n_d$** and **domain size $N_d$**, if known.

Direct estimates are paired with estimated **sampling variances**, telling us how much uncertainty comes from observing only a sample within each domain. In practice, the variance depends on the estimator, weight scale, and survey design information available.

## Next Step

Use the direct estimates and sampling variances to decide whether direct reporting is enough, or whether model-based SAE is needed.

| Scenario | Next step |
|---|---|
| Most domains have usable sample sizes and acceptable uncertainty | Direct estimates may be enough for reporting. |
| Many domains have no sample, or sampling variances are missing, zero, or unreasonable | Revisit the domain choice, estimand, or input data before modelling. |
| Direct estimates are imprecise and a quick model-based approach is enough | Fit a [Frequentist Area-Level SAE](freq-area-sae.qmd) model to borrow strength from covariates. |
| Direct estimates are imprecise, spatial effects matter, or uncertainty needs to be reported as probabilities | Fit a [Bayesian Area-Level SAE with Spatial](bayesian-area-sae.qmd) model to borrow strength from covariates and neighbouring areas. |

## Code Implementation

:::{.callout-tip}
Use [`sae::direct()`](https://search.r-project.org/CRAN/refmans/sae/html/direct.html) to calculate direct estimates by domain. It selects an estimation method automatically based on the provided inputs: 

- Horvitz-Thompson estimator: pass `sweight` and pre-aggregated `domsize`.
- Hajek ratio estimator: pass `sweight`, and `domsize` equals sum of weights as pseudo-domain-size.
- If neither is provided, it defaults to unweighted sample mean.
:::

```{r}
# Download pre-generated analysis-ready files into a temp folder
library(dplyr)
library(readr)
library(stringr)
library(tidyr)
library(sae)
library(sf)
library(ggplot2)
library(patchwork)

# Load analysis-ready data
zip_tmp <- tempfile(fileext = ".zip")
url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-clean-data.zip"
download.file(url, zip_tmp, mode = "wb")
unzip(zip_tmp, exdir = tempdir())
base_path <- file.path(tempdir(), "clean")

# Load shapefiles for plotting
boundary_zip <- tempfile(fileext = ".zip")
boundary_url <- "https://raw.githubusercontent.com/shaunhoang/tutorial-site/main/data/tutorial-boundaries.zip"
download.file(boundary_url, boundary_zip, mode = "wb")
unzip(boundary_zip, exdir = tempdir())

msoa_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-msoa.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = MSOA21CD) |>
  filter(str_starts(domain_id, "W"))

la_boundaries <- st_read(
  file.path(tempdir(), "boundaries", "boundaries-ltla.geojson"),
  quiet = TRUE
) |>
  rename(domain_id = LAD21CD) |>
  filter(str_starts(domain_id, "W"))

# Create master lists of MSOAs and LAs
msoa_master_list <- read_csv(
  file.path(base_path, "domain_sizes_adult_msoa.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_name) |>
  distinct()

la_master_list <- read_csv(
  file.path(base_path, "domain_sizes_population_la.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_name) |>
  distinct()
```

### Direct Estimation
We will perform Direct Estimation for each of the three indicators differently according to data availability

- `commute_distance`: Hajék estimator (no known domain sizes)
- `poor_bus_satisfaction`: Hórvitz-Thompson estimator
- `walk_weekly`: Hórvitz-Thompson estimator

#### Mean Commute Distance
Load the `commute_distance` survey indicator. It has no known worker domain-size file, so the sum of analysis weights in each sampled MSOA is used as a pseudo domain size since `sae::direct()` requires a `domsize`.
```{r}
survey_indicator <- read_csv(
  file.path(base_path, "ind_weight_commute_distance.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id))

domain_sizes_pseudo <- survey_indicator |>
  group_by(domain_id) |>
  summarise(domain_size = sum(weight), .groups = "drop") |>
  as.data.frame()

dir_est_commute <- sae::direct(
  y = survey_indicator$indicator,
  dom = survey_indicator$domain_id,
  sweight = survey_indicator$weight,
  domsize = domain_sizes_pseudo,
  replace = TRUE 
  # required for sae::direct() to calculate variance when using Hajek estimator         
)

dir_est_commute$Var <- dir_est_commute$SD^2
dir_est_commute$Lower <- dir_est_commute$Direct - 1.96 * dir_est_commute$SD
dir_est_commute$Upper <- dir_est_commute$Direct + 1.96 * dir_est_commute$SD
head(dir_est_commute)
```

#### Dissatisfaction with bus service
Load the `poor_bus_satisfaction` survey indicator and its matching adult MSOA domain-size file.
```{r}
survey_indicator_bus <- read_csv(
  file.path(base_path, "ind_weight_poor_bus_satisfaction.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id))

msoa_adult_domain_sizes <- read_csv(
  file.path(base_path, "domain_sizes_adult_msoa.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_size) |>
  as.data.frame()

dir_est_bus <- sae::direct(
  y = survey_indicator_bus$indicator,
  dom = survey_indicator_bus$domain_id,
  sweight = survey_indicator_bus$weight,
  domsize = msoa_adult_domain_sizes
)

dir_est_bus$Var <- dir_est_bus$SD^2
dir_est_bus$Lower <- dir_est_bus$Direct - 1.96 * dir_est_bus$SD
dir_est_bus$Upper <- dir_est_bus$Direct + 1.96 * dir_est_bus$SD

head(dir_est_bus)
```

#### Walking
Load the `walk_weekly` survey indicator and its matching total population LA domain-size file.
```{r}
survey_indicator_walk <- read_csv(
  file.path(base_path, "ind_weight_walk_weekly.csv"),
  show_col_types = FALSE
) |>
  filter(!is.na(indicator), !is.na(weight), !is.na(domain_id))

domain_sizes_walk <- read_csv(
  file.path(base_path, "domain_sizes_population_la.csv"),
  show_col_types = FALSE
) |>
  dplyr::select(domain_id, domain_size) |>
  as.data.frame()

dir_est_walk <- sae::direct(
  y = survey_indicator_walk$indicator,
  dom = survey_indicator_walk$domain_id,
  sweight = survey_indicator_walk$weight,
  domsize = domain_sizes_walk
)

dir_est_walk$Var <- dir_est_walk$SD^2
dir_est_walk$Lower <- dir_est_walk$Direct - 1.96 * dir_est_walk$SD
dir_est_walk$Upper <- dir_est_walk$Direct + 1.96 * dir_est_walk$SD

head(dir_est_walk)
```

### Evaluate Direct Estimates

The key questions now are whether any domains have no direct estimates or variance, and whether the direct estimates that do exist are precise enough to report. The two metrics to pay attention to are:

- `median_cv`: the typical coefficient of variation, calculated as standard error divided by the estimate. If above 100%, it means the typical standard error is larger than the estimate itself. 
- `median_ci_width`: the typical width of the 95% confidence interval in real terms.

```{r}
#| echo: false
summarise_direct_estimates <- function(result, master_domains) {
  domains_no_data <- master_domains |>
    anti_join(
      result |> transmute(domain_id = Domain),
      by = "domain_id"
    )
  
  result |>
    summarise(
      target_domain_n = nrow(master_domains),
      domains_with_direct_estimate = n(),
      domains_with_no_data = nrow(domains_no_data),
      domains_with_missing_variance = sum(is.na(Var) | is.nan(Var)),
      median_sample_size = median(SampSize, na.rm = TRUE),
      median_cv = median(CV, na.rm = TRUE),
      median_ci_width = median(Upper - Lower, na.rm = TRUE)
    ) |>
    pivot_longer(
      cols = everything(),
      names_to = "diagnostic",
      values_to = "value"
    )
}

plot_direct_estimates <- function(result, boundaries, title, border_colour = NA) {
  map_data <- boundaries |>
    left_join(
      result,
      by = c("domain_id" = "Domain")
    )
  
  estimate_map <- ggplot(map_data) +
    geom_sf(aes(fill = Direct), colour = border_colour, linewidth = 0.1) +
    scale_fill_distiller(palette = "YlGnBu", direction = 1, na.value = "grey85") +
    labs(
      fill = "Estimate",
      title = "Direct estimate"
    ) +
    theme_void()
  
  cv_map <- ggplot(map_data) +
    geom_sf(aes(fill = CV), colour = border_colour, linewidth = 0.1) +
    scale_fill_distiller(palette = "OrRd", direction = 1, na.value = "grey85") +
    labs(
      fill = "CV",
      title = "Coefficient of variation"
    ) +
    theme_void()
  
  estimate_map + cv_map +
    plot_annotation(title = title)
}
```

- **Mean Commute Distance**

```{r}
#| echo: false
print(summarise_direct_estimates(dir_est_commute, msoa_master_list))
```

```{r}
#| echo: false
print(
  plot_direct_estimates(
    dir_est_commute,
    msoa_boundaries,
    "MSOA commute_distance direct estimate and coefficient of variation"
  )
)
```

- **Dissatisfaction with bus service**

```{r}
#| echo: false
print(summarise_direct_estimates(dir_est_bus, msoa_master_list))
```

```{r}
#| echo: false
print(
  plot_direct_estimates(
    dir_est_bus,
    msoa_boundaries,
    "MSOA poor_bus_satisfaction direct estimate and coefficient of variation"
  )
)
```

- **Walking**

```{r}
#| echo: false
print(summarise_direct_estimates(dir_est_walk, la_master_list))
```

```{r}
#| echo: false
print(
  plot_direct_estimates(
    dir_est_walk,
    la_boundaries,
    "LA walk_weekly direct estimate and coefficient of variation",
    border_colour = "white"
  )
)
```
### Final thoughts

Overall, the direct estimates are useful as inputs, but model-based estimation is likely needed to produce more stable small-area estimates.

That said, since model-based estimation methods work on these direct estimates and variances, having a solid base is essential before moving forward.

In general, the high the sample size per domain, the more stable the estimates. Moreover, when domain sizes are known, the precision is also higher (`walk-weekly` and `poor_bus_satisfaction`'s median CVs are below 100%). By contrast, the estimates for `commute_distance` rely only on weights and has much higher relative uncertainty. This illustrates why suitable domain-size information is valuable in SAE.

In the last scenario, another remedy is to redefine the estimand associated with `commute_distance` so that it applies to a population with known domain sizes, such as all adults, rather than a subgroup whose MSOA totals may be unavailable. However, this would essentially change what is being measured.
