---
title: "Bayesian Area and Spatial Smoothing"
---

Spatial SAE adds structure when neighbouring domains are expected to have similar values after accounting for covariates. Bayesian SAE extends this by representing uncertainty through posterior distributions.

## When spatial structure helps

Spatial modelling is most defensible when:

-   The indicator has a plausible spatial process.
-   Domains are geographically meaningful for the process.
-   Residuals from a non-spatial model show spatial patterning.
-   The adjacency structure is reliable and aligned with the modelling table.

## When to be cautious

Spatial smoothing can hide real local differences. It can also create overconfident maps if adjacency is treated as a substitute for missing survey signal.

## Adjacency checks

```{r}
library(sf)
library(spdep)

domains_sf <- st_read("../data/boundaries/boundaries-ltla.geojson")

neighbors <- poly2nb(domains_sf, queen = TRUE)
adjacency <- nb2mat(neighbors, style = "B", zero.policy = TRUE)

stopifnot(nrow(adjacency) == nrow(domains_sf))
```

## Bayesian FH with spatial effects

This section should introduce the practical pattern for `SUMMER` and `INLA`:

1.  Prepare direct estimates and variances.
2.  Prepare domain covariates.
3.  Prepare an adjacency graph.
4.  Fit a non-spatial Bayesian FH model.
5.  Fit a spatial FH model.
6.  Compare posterior summaries, residuals, and mapped uncertainty.

```{r}
# Placeholder pattern only. Final code will depend on the selected package API.

# fit <- SUMMER::smoothSurvey(
#   data = direct_estimates,
#   Amat = adjacency,
#   responseType = "gaussian",
#   formula = direct ~ covariate_1 + covariate_2
# )
```

## Outputs to communicate

-   Posterior mean or median.
-   Credible intervals.
-   Exceedance probabilities for policy thresholds.
-   Maps of uncertainty, not only maps of point estimates.